\bf Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution

### 2024-07-05 paper.qmd
abstract and affiliation

---
title: "Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution"
author: "Annalisa Calvi, Ursula Laa, Dianne Cook"
format:
  jasa-pdf:
    keep-tex: true    
bibliography: bibliography.bib  
header-includes: | 
  \usepackage{amsmath}
  \usepackage{amsthm}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\kD{$k\text{-}D$}
  \newcommand\dD{$d\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
  \newtheorem*{theorem}{Theorem}
---

@@

## Projecting an ellipsoid

@@

![Random projection.](images/anomaly1.png){#fig-anomaly1 width=300 align="center"}

![Optimal projection.](images/anomaly2.png){#fig-anomaly2 width=300 align="center"}

@@

![Sample of female patients.](images/liver1.png){#fig-liver1 width=300 align="center"}

![Longitudinal profile of an aging patient.](images/liver2.png){#fig-liver2 width=300 align="center"}

>>>>

---
title: "Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution"
author: 
  - name: Annalisa Calvi
    affiliations:
      name: Monash University
      department: School of Mathematics
  - name: Ursula Laa
    affiliations:
      name: University of Natural Resources and Life Sciences, Vienna
      department: Institute of Statistics
    email: ursula.laa@boku.ac.at
    ORCID: 0000-0002-0249-6439
  - name: Dianne Cook
    affiliations:
      name: Monash University
      department: Department of Econometrics and Business Statistics
    email: dicook@monash.edu
    ORCID: 0000-0002-3813-7155
format:
  jasa-pdf:
    keep-tex: true    
abstract: |
  Many data problems contain some reference or normal conditions, upon which to compare newly collected data. This scenario occurs in data collected as part of clinical trials to detect adverse events, and or for measuring climate change against historical norms. The data is typically multivariate, and often the normal ranges are specified by a multivariate normal distribution. The work presented in this paper develops methods to compare the new sample against the reference distribution with high-dimensional visualisation. It uses a projection pursuit guided tour to produce a sequence of low-dimensional projections steered towards those where the new sample is most different from the reference. A new projection pursuit index is defined, and the drawing of the projected ellipse is computed analytically. The methods are implemented in the R package, `tourr`. 
keywords:
  - exploratory data analysis
  - statistical graphics
  - dimension reduction
  - high-dimensional visualisation
  - climate change
  - clinical trials
bibliography: bibliography.bib  
header-includes: | 
  \usepackage{amsmath}
  \usepackage{amsthm}
  \usepackage{float}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{bm}
  \def\tightlist{}
  \usepackage{setspace}
  \newcommand\pD{$p\text{-}D$}
  \newcommand\kD{$k\text{-}D$}
  \newcommand\dD{$d\text{-}D$}
  \newcommand\gD{$2\text{-}D$}
  \newtheorem*{theorem}{Theorem}
---

@@

### Projecting an ellipsoid

@@

![Random projection](images/anomaly1.png){#fig-anomaly1 width=300 align="center"}

![Optimal projection](images/anomaly2.png){#fig-anomaly2 width=300 align="center"}

@@

![Sample of female patients](images/liver1.png){#fig-liver1 width=300 align="center"}

![Longitudinal profile of an aging patient](images/liver2.png){#fig-liver2 width=300 align="center"}

### 2024-07-05 paper.tex
abstract and affiliation

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution},
  pdfauthor={Annalisa Calvi, Ursula Laa, Dianne Cook},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\title{\bf Is this normal? A new projection pursuit index to assess a
sample against a multivariate null distribution}
\author{
Annalisa Calvi, Ursula Laa, Dianne Cook\\
}
\maketitle

\bigskip
\bigskip
\begin{abstract}

\end{abstract}

@@

\section{Projecting an ellipsoid}\label{projecting-an-ellipsoid}

@@

\subcaption{\label{fig-anomaly1}Random projection.}

@@

\subcaption{\label{fig-anomaly2}Optimal projection.}

@@

\subcaption{\label{fig-liver1}Sample of female patients.}

@@

\subcaption{\label{fig-liver2}Longitudinal profile of an aging patient.}

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution},
  pdfauthor={Annalisa Calvi; Ursula Laa; Dianne Cook},
  pdfkeywords={exploratory data analysis, statistical
graphics, dimension reduction, high-dimensional visualisation, climate
change, clinical trials},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\title{\bf Is this normal? A new projection pursuit index to assess a
sample against a multivariate null distribution}
\author{
Annalisa Calvi\\
School of Mathematics, Monash University\\
and\\Ursula Laa\\
Institute of Statistics, University of Natural Resources and Life
Sciences, Vienna\\
and\\Dianne Cook\\
Department of Econometrics and Business Statistics, Monash University\\
}
\maketitle

\bigskip
\bigskip
\begin{abstract}
Many data problems contain some reference or normal conditions, upon
which to compare newly collected data. This scenario occurs in data
collected as part of clinical trials to detect adverse events, and or
for measuring climate change against historical norms. The data is
typically multivariate, and often the normal ranges are specified by a
multivariate normal distribution. The work presented in this paper
develops methods to compare the new sample against the reference
distribution with high-dimensional visualisation. It uses a projection
pursuit guided tour to produce a sequence of low-dimensional projections
steered towards those where the new sample is most different from the
reference. A new projection pursuit index is defined, and the drawing of
the projected ellipse is computed analytically. The methods are
implemented in the R package, \texttt{tourr}.
\end{abstract}

\noindent%
{\it Keywords:} exploratory data analysis, statistical
graphics, dimension reduction, high-dimensional visualisation, climate
change, clinical trials
\vfill

@@

\subsection{Projecting an ellipsoid}\label{projecting-an-ellipsoid}

@@

\subcaption{\label{fig-anomaly1}Random projection}

@@

\subcaption{\label{fig-anomaly2}Optimal projection}

@@

\subcaption{\label{fig-liver1}Sample of female patients}

@@

\subcaption{\label{fig-liver2}Longitudinal profile of an aging patient}
